---
title: "Statistical Modeling of Calorie Burn in Gym Workouts"
author: "Natali Kostadinovic"
date: "2024-12-02"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    number_sections: true
    fig_caption: true
editor_options: 
  markdown: 
    wrap: 72
---

**1.Introduction:**

Understanding how personal and workout-related factors affect the number
of calories burned during exercise can help design personalized fitness
and nutrition programs. This project uses a dataset of gym members to
explore the relationship between calorie expenditure and key predictors
including age, gender, workout type, session duration, and body fat
percentage.

The main research question is:

> **Which variables significantly influence the number of calories
> burned in a workout session?**

The analysis includes data exploration, regression modeling, outlier
diagnostics, and model selection using R.

The dataset, originally published on Kaggle, is simulated but based on
realistic workout patterns and public health data.

```{r, out.width="50%", warning=FALSE, include = T}
library(ggplot2)
library(dplyr)
library(car)
library(readr)
library(leaps)
library(gridExtra)
library(GGally)
gym <- read_csv("gym_members_exercise_tracking.csv")
names(gym) <- c("Age", "Gender", "Weight", "Height", "Max_BPM", "Avg_BPM", "Resting_BPM", "Session_Duration", "Calories_Burned", "Workout_Type", "Fat_Percentage", "Water_Intake", "Workout_Frequency", "Experience_Level", "BMI")
set.seed(123)
names(gym)
ggpairs(gym %>% select(Calories_Burned, Age, Session_Duration, Water_Intake, Fat_Percentage, Workout_Type, Gender))


#Source: https://www.kaggle.com/datasets/valakhorasani/gym-members-exercise-dataset?fbclid=IwY2xjawGQqDNleHRuA2FlbQIxMAABHWXxbjoBhN60-umXAWeqGOAXQBUGJm_MvQbBSIrhAGK3GzQFExUwsmU3Cg_aem_ON8I_GTOV_zi3KofVSCWxA
```

```{r}
lm_full_before<-lm(Calories_Burned~Age+Session_Duration+Fat_Percentage+Gender + Workout_Type,
 data=gym)
 gym<-gym%>%
 mutate(
 resid=residuals(lm_full_before)
 )
 
ggplot(data=gym,mapping=aes(x=resid))+
 geom_density()+
 ggtitle("Density plot of Residuals before Transformation")
 theme_bw()

```

```{r}
summary_stats1 <- gym %>%
summarize(
  count = n(),            # Count of rows
  mean_Age = mean(Age), # Mean of the column
  sd_Age = sd(Age)      # Standard deviation
)
print(summary_stats1)
summary_stats2 <- gym %>%
summarize(
  count = n(),            # Count of rows
  mean_Session_Duration = mean(Session_Duration), # Mean of the column
  sd_Session_Duration = sd(Session_Duration)      # Standard deviation
)
print(summary_stats2)

summary_stats <- gym %>%
summarize(
  count = n(),            # Count of rows
  mean_Fat_Percentage = mean(Fat_Percentage), # Mean of the column
  sd_Fat_Percentage = sd(Fat_Percentage)      # Standard deviation
)
print(summary_stats)

```

```{r}
ggpairs(gym%>%select(Age,Session_Duration,Workout_Type,Fat_Percentage, Gender, Calories_Burned))
```

**1. Model assumptions diagnostics [ Variable "Age"]:**

-   Independence: We cannot determine independence without additional
    information on how the data was collected; however, we will assume
    it is independent.

-   Linearity: Age vs. Calories_Burned: From plot 1 and 2, we can say
    that the relationship is linear.

-   Equal Variance: The variance assumption is not satisfied at this
    step, as seen in the Response vs. Explanatory (plot 1) and Residuals
    vs. Explanatory.(plot 2)

-   Normality: From plot 3, we can say that it is pretty much normally
    distributed.

-   Outliers: From the Response vs. Explanatory (plot 1) and Residuals
    vs. Explanatory (plot 2), we observe that there are no outliers.

```{r}
gym_transformed<-gym%>%
 transmute(
   Age = Age,
   log_Session_Duration =log(Session_Duration),
   Workout_Type= Workout_Type,
   sqrt_Fat_Percentage=sqrt(Fat_Percentage),
   Gender= Gender,
   log_Calories_Burned=log(Calories_Burned)
 )
ggpairs(gym_transformed)



```

**Interaction Terms Detection:**

```{r}
ggplot(data=gym_transformed,aes(x=Age,y=log_Calories_Burned,color=Workout_Type))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()

ggplot(data=gym_transformed,aes(x=sqrt_Fat_Percentage,y=log_Calories_Burned,color=Workout_Type))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()

ggplot(data=gym_transformed,aes(x=log_Session_Duration,y=log_Calories_Burned,color=Workout_Type))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()

ggplot(data=gym_transformed,aes(x=Age,y=log_Calories_Burned,color=Gender))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()

ggplot(data=gym_transformed,aes(x=sqrt_Fat_Percentage,y=log_Calories_Burned,color=Gender))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()

ggplot(data=gym_transformed,aes(x=log_Session_Duration,y=log_Calories_Burned,color=Gender))+
 geom_point()+
 geom_smooth(method="lm",se=FALSE)+
 theme_bw()




```

While non-parallel slopes in exploratory plots suggested possible
interaction effects, we formally tested interaction terms in extended
models. These terms were not statistically significant and did not
improve model fit based on BIC. Therefore, we proceeded without
interaction terms for clarity and interpretability.

```{r}
lm_full<-lm(log_Calories_Burned~Age+log_Session_Duration+sqrt_Fat_Percentage+Gender + Workout_Type,
 data=gym_transformed)
 gym_transformed<-gym_transformed%>%
 mutate(
 resid=residuals(lm_full)
 )
 p1<-ggplot(data=gym_transformed,mapping=aes(x=Age,y=resid))+
 geom_point()+
 geom_smooth()+
 theme_bw()
 p2<-ggplot(data=gym_transformed,mapping=aes(x=log_Session_Duration,y=resid))+
 geom_point()+
 geom_smooth()+
 theme_bw()
 p3<-ggplot(data=gym_transformed,mapping=aes(x=Workout_Type,y=resid))+
 geom_point()+
 geom_smooth()+
 theme_bw()
 p4<-ggplot(data=gym_transformed,mapping=aes(x=sqrt_Fat_Percentage,y=resid))+
 geom_point()+
 geom_smooth()+
 theme_bw()
 p5<-ggplot(data=gym_transformed,mapping=aes(x=Gender,y=resid))+
 geom_point()+
 geom_smooth()+
 theme_bw()
 p6<-ggplot(data=gym_transformed,mapping=aes(x=resid))+
 geom_density()+
 theme_bw()
 grid.arrange(p1,p2,p3,p4,p5,p6)
```

```{r}
summary(lm_full)
```

**2. Outlier diagnostics**

```{r}
gym_transformed<-gym_transformed%>%
 mutate(
 obs_index=row_number(),
 h=hatvalues(lm_full),
 studres=rstudent(lm_full),
 D=cooks.distance(lm_full)
 )
 ggplot(data =gym_transformed, mapping=aes(x=obs_index,y=h))+
 geom_hline(yintercept=2*8/nrow(gym_transformed))+ # 8 is the number of betas
 geom_point()+
 theme_bw()
 
ggplot(data=gym_transformed,mapping=aes(x=obs_index,y=studres))+
 geom_point()+
 theme_bw()

ggplot(data=gym_transformed,mapping=aes(x=obs_index,y=D))+
 geom_point()+
 theme_bw()
```

```{r}
leverage_indices <- which(gym_transformed$h > 2*8/nrow(gym_transformed))
cooks_indices <- which(gym_transformed$D > 0.005)
studres_indices <- which((gym_transformed$studres > 3|gym_transformed$studres < -3))
obs_to_investigate <- unique(c(leverage_indices, cooks_indices, studres_indices))
obs_to_investigate
```

```{r}
gym_transformed[obs_to_investigate, ]
```

```{r}
round(summary(lm_full)[4] $coefficients,4)
```

```{r}
gym_transformed_minus_sus <-gym_transformed[-obs_to_investigate, ]
lm_fit_without_suspicious <- lm(log_Calories_Burned~Age+log_Session_Duration+sqrt_Fat_Percentage+Gender + Workout_Type, data = gym_transformed_minus_sus)
round(summary(lm_fit_without_suspicious,digits=3)[4]$coefficients,4)

```

Determine whether outliers are influential or not:

-    Change in values in Beta's:

    -   Age: within 2 SEs of original estimate

    -   log_Session_Duration: within 2 SEs of original estimate

    -   sqrt_Fat_Percentage: within 2 SEs of original estimate

    -   Workout_TypeHIIT: within 2 SEs of original estimate

    -   Workout_TypeStrength: within 2 SEs of original estimate

    -   Workout_TypeYoga: within 2 SEs of original estimate

-   Change in signs:

    -   Workout_TypeStrength: from 0.0011 to -0.00063

-   Change in p-value:

    -   No significant change in p-values

**Conclusion: The identified outliers do not have a significant
influence on the analysis.**

**Model Selection**

```{r}
gym_models <- regsubsets(log_Calories_Burned~Age+log_Session_Duration+sqrt_Fat_Percentage+Gender + Workout_Type, data=gym_transformed)
plot(gym_models)
```

```{r}
# Variables included in the models
summary(gym_models)[[1]]*1

# BIC for four models
summary(gym_models)$bic
```

```{r}
# Plot the BIC
bic <- data.frame(Model = 1:7, BIC = summary(gym_models)$bic)
 ggplot(data=bic, aes(x=Model, y=BIC)) +
 geom_point(aes(color=BIC < -2190), size=3) +
 geom_line() +
 theme_bw() +
 theme(legend.position = "none")
```

-   Model 3, Model 4, and Model 5 have roughly similar performance:

    -   Model 3: Age, log_Session_Duration, GenderMale (versus
        GenderFemale) (BIC=-2206.404)

    -   Model 4: Age, log_Session_Duration, GenderMale (versus
        GenderFemale), Workout_TypeYoga (versus other
        Workout_TypenotYoga) (BIC=-2200.778)

    -   Model 5: Age, log_Session_Duration, GenderMale (versus
        GenderFemale), Workout_TypeYoga and Workout_TypeHIIT (versus
        Workout_TypeStrength) (BIC=-2194.729)

```{r}
# Effect of the outliers on model selection
gym_models2 <- regsubsets(log_Calories_Burned~Age+log_Session_Duration+sqrt_Fat_Percentage+Gender + Workout_Type, data = gym_transformed_minus_sus)
plot(gym_models2)
 
```

```{r}
#Variables included in the models
summary(gym_models2)[[1]]*1
# BIC for seven models
summary(gym_models2)$bic
```

-   Model 3', Model 4', and Model 5' have roughly similar performance:

    -   Model 3': Age, log_Session_Duration, GenderMale (versus
        GenderFemale) (BIC=-2206.404)

    -   Model 4': Age, log_Session_Duration, GenderMale (versus
        GenderFemale), Workout_TypeYoga (versus Workout_TypenotYoga)
        (BIC=-2200.778)

    -   Model 5': Age, log_Session_Duration, GenderMale (versus
        GenderFemale), Workout_TypeYoga and Workout_TypeHIIT (versus
        Workout_TypenotYogaHIIT) (BIC=-2194.729)

## Check if outliers have an effect on the model selection.

# A. Model selection

-   No difference in model 3 and 3', 4 and 4', 5 and 5'

# B. Differences in the model fit:

```{r}
# Create new variables Workout_TypenotYoga
#All data
gym_transformed <- gym_transformed %>%
 mutate(
 Workout_Type_reduced_Yoga = ifelse(Workout_Type %in% c("HIIT", "Strength"), "other", "Yoga"),
 Workout_Type_reduced_YogaHIIT = ifelse(Workout_Type %in% c("HIIT", "Strength"), "other", Workout_Type)
 )

# Without outliers
gym_transformed_minus_sus<- gym_transformed_minus_sus %>%
 mutate(
 Workout_Type_reduced_Yoga = ifelse(Workout_Type %in% c("HIIT", "Strength"), "other", "Yoga"),
 Workout_Type_reduced_YogaHIIT = ifelse(Workout_Type %in% c("HIIT", "Strength"), "other", Workout_Type)
 )
 
# Fit models 3, 4 and 5 with all the data
fit3_all <- lm(log_Calories_Burned~Age+log_Session_Duration+Gender, data=gym_transformed)
fit4_all <- lm(log_Calories_Burned~Age+log_Session_Duration+Gender+Workout_Type_reduced_Yoga, data=gym_transformed)
fit5_all <- lm(log_Calories_Burned~Age+log_Session_Duration+Gender+Workout_Type_reduced_YogaHIIT, data=gym_transformed)
 
 # Fit models 2, 3, and 4 without the outliers
fit3_minus_suspicious <- lm(log_Calories_Burned~Age+log_Session_Duration+Gender, data=gym_transformed_minus_sus)
fit4_minus_suspicious <- lm(log_Calories_Burned~Age+log_Session_Duration+Gender+Workout_Type_reduced_Yoga, data=gym_transformed_minus_sus)
fit5_minus_suspicious <- lm(log_Calories_Burned~Age+Age+log_Session_Duration+Gender+Workout_Type_reduced_YogaHIIT, data=gym_transformed_minus_sus)

summary(fit3_all)$coefficients
summary(fit3_minus_suspicious)$coefficients
summary(fit4_all)$coefficients
summary(fit4_minus_suspicious)$coefficients
summary(fit5_all)$coefficients
summary(fit5_minus_suspicious)$coefficients
```

**No significant change in values or p-values with and without
outliers.**

**Model statement:**

Based on the model selection criteria (BIC) and diagnostics, the best
model includes `Age`, `log(Session_Duration)`, and `Gender`. The fitted
regression equation is:

$$
\hat{\mu}_{\log(\text{Calories_Burned})} = 6.653 + 0.0034 \cdot \text{Age} + 1.0039 \cdot \log(\text{Session Duration}) + 0.098 \cdot \text{GenderMale}
$$

Where: - `GenderMale = 1` if the subject is male, and `0` if female.

## **Multicollinearity.**

```{r}
vif(fit3_all)
vif(fit3_minus_suspicious)

```

There's no multicollinearity issue with our data and selected models
since all VIFs are below 4.

**Hypothesis Testing**

Let the response variable be `log(Calories_Burned)`. We test the
following:

-   **Age:**\
    $H_0: \beta_1 = 0$ \
    $H_A: \beta_1 \neq 0$

-   **Session Duration:**\
    $H_0: \beta_2 = 0$\
    $H_A: \beta_2 \neq 0$

-   **Gender:**\
    $H_0: \beta_3 = 0$\
    $H_A: \beta_3 \neq 0$

We use the 5% significance level for all tests.

### Hypothesis Test: Age

From the summary of the final model (Model 3), the coefficient for `Age`
is:

-   p-value: **\< 2e-16**

#### Interpretation:

Because the p-value is extremely small, we have very strong statistical
evidence against the null hypothesis. Therefore, we reject the null
hypothesis and conclude that age has a statistically significant effect
on the log of calories burned.

#### Conclusion:

There is **very strong evidence** that age has a statistically
significant effect on the log of calories burned, holding other
variables constant. Specifically, for each additional year of age, the
**log of calories burned decreases by approximately 0.0034 units**, on
average.

### Hypothesis Test: Session Duration

From the model summary, the coefficient for log(Session_Duration) is:

p-value: **\< 2e-16**

#### Interpretation:

Since the p-value is much smaller than 0.05 (virtually 0), the observed
effect is highly unlikely to have occurred by random chance under the
null. We reject the null hypothesis and conclude that session duration
significantly affects the log of calories burned.

#### Conclusion:

There is extremely strong evidence that session duration has a
statistically significant positive effect on the log of calories burned.
Specifically, a 1-unit increase in log(session duration) is associated
with an increase of approximately 1.0039 units in the log of calories
burned, on average.

### Hypothesis Test: Gender

From the model output, the coefficient for GenderMale (compared to
female) is:

p-value: **\< 2e-45**

### Interpretation:

Given that the p-value (3.98 × 10⁻⁴⁵) is far less than 0.05, we find
overwhelming evidence that gender influences calorie expenditure. We
reject the null hypothesis and conclude that gender has a significant
effect on the log of calories burned.

### Conclusion:

There is extremely strong statistical evidence that gender affects the
log of calories burned. On average, males burn approximately 0.098 more
log-calories than females, holding age and session duration constant.

## Model Selection and Final Model Justification

Using `regsubsets()` and BIC as the model selection criterion, we
compared several candidate models that included different combinations
of predictors. The three best models (Model 3, 4, and 5) had similar BIC
values, but:

-   **Model 3** (with Age, log(Session_Duration), and Gender) had the
    **lowest BIC = -2206.404**, indicating the best balance between
    model fit and complexity.
-   Models 4 and 5 added additional levels of `Workout_Type`, but the
    BIC improvement was not substantial, and their coefficients were
    statistically insignificant.

Additionally, multicollinearity was assessed using the Variance
Inflation Factor (VIF). All VIFs for Model 3 were near 1 (well below the
common threshold of 4), both **with and without outliers**.

Therefore, **Model 3 is selected as the final model** based on: - Best
BIC score - Interpretability - Lack of multicollinearity - Robustness to
outliers

------------------------------------------------------------------------

**Final Model Statement and Interpretation**

$$
\hat{\mu}_{\log(\text{Calories_Burned})} = 6.653 + 0.0034 \cdot \text{Age} + 1.0039 \cdot \log(\text{Session Duration}) + 0.098 \cdot \text{GenderMale}
$$

Where: - `GenderMale = 1` if male, `0` if female.

**This model suggests that:**

-   Older age slightly **decreases** log-calories burned (β = -0.0034)

-   Longer session duration significantly **increases** it (β =
    1.0039) - Males, on average, burn **more** than females (β = 0.098),
    controlling for other variables.

**Conclusion:**

-   Age negatively affects log(calories burned)

-   Longer workouts significantly increase calorie expenditure

-   Males tend to burn more calories than females

-   No significant impact from workout type after adjusting for others
